Giant Low Surface Brightness Galaxies As Merger Remnants

### 2016-07-27 glsb.tex
add ref
edits

\altaffiltext{1}{Department of Astronomy, University of Washington, Box 351580,
Seattle WA, 98195}
\altaffiltext{2}{XXX-U Mich}
\altaffiltext{3}{Jeremiah Horrocks Institute, University of Central Lancashire, Preston PR1 2HE, UK.  RCUK Fellow}
\altaffiltext{4}{XXX-Cal Tech}

\begin{abstract}
We present IFU observations of the giant low surface brightness galaxies Malin~2 and UGC~6614.  We fit the stellar dynamics of these systems and compare their spatially resolved spectra to star formation history models.  We argue that Malin~2 and UGC~6614 are normal elliptical galaxies which have undergone a high angular momentum mergers which formed their GLSB disks, making them unique systems for probing the dark matter halos of elliptical systems.  We find the stellar dynamics are consistent with earlier measurements of the GLSB \HI\ disks.  The stellar populations of the GLSB disks are dominated by old metal-poor stars, while the central ellipticals are metal rich.

@@

Observers tend to classify GLSB systems as normal spirals.  \citet{Matthews01b} list Malin~2 as an Scd while \citet{Das13} concludes that GLSB systems are an extreme form of late type spiral galaxies.  Similarly, \citet{Kasparova14} claim that the large dark matter scale of Malin~2 can explain it's features.  Even the Galaxy Zoo project classifies GLSB galaxies as spirals.  While GLSB systems superficially resemble normal spiral galaxies, they are clearly extreme outliers when their scale lengths are compared to other galaxy samples, being around a factor of 2 larger than the largest typical spirals \citep{Dalcanton97}.

The disks in GLSB systems are incredibly large, with Malin~1 being detected in $R$-band to a radius of over 100 kpc \citep{Moore06}.  Disks of this size are a challenge for typical $\Lambda$CDM cosmology galaxy formation and exotic formation scenarios have been proposed for GLSB systems.  \citet{Mapelli08} form GLSB-like galaxies in simulations via the fading of collisional ring galaxies.  In their model, a collision sets off an expanding ring of star formation.  As this the star formation subsides and the stellar population ages, the ring galaxy fades over $\sim$100 Myr to resembles observed GLSBs.  One obvious problem with this scenario is that most GLSB's are relatively isolated, and thus there are no obvious candidates for the impactor systems that would have triggered the star formation burst.

The another formation scenario for GLSB systems is that they are normal galaxies that undergo a high angular momentum minor merger where the LSB disk is the shredded remains of an accreted galaxy.  \citet{Pen06} find that extended disks can be formed via the disruption of dwarf galaxies.  These simulated debris disks have exponential profiles and extend out to 200 kpc from the central galaxy.

@@

In comparison, the stellar disk of Malin~2 extends beyond 40 kpc (5.9 $R_e$), and the HI disk reaches beyond 100 kpc (14.8 $R_e$).  \citet{Galaz15} take deep images of Malin~1 and find the stellar disk extends to 160 kpc.  These distances have occasionally been reached by x-ray observations \citep{Buote12}.

@@

In one of the most extensive studies of a gLSB system, \citet{Kasparova14} combine ARC 3.5m photometry and long-slit spectra with Gemini spectra to mass-model Malin~2.  They find the broadband colors are not consistent with a catastrophic formation, but are best fit by exponentially declining SFH.  They speculate the main difference between Malin~2 and non-giant LSB galaxies is the dark matter halo scale.  To some extent, this is not a particularly satisfying conclusion because one is then left with the mystery of why Malin~2's dark matter halo is so different from non-giant galaxies.

@@

\section{Observations}

@@

For each galaxy, observations were taken at three unique dither positions to provide nearly complete spatial coverage.  Individual science exposures were 30 minutes.  Malin~2 had 10 exposures per dither position (11 exposures for dither position 1), for a total of 15 hours of exposure time.  Observations of UGC 06614 had alignment issues between observing runs.  This resulted in a deep stack of 6 exposures, and a shallow stack with only 2 exposures.  With only 2 exposures, cosmic ray rejection becomes difficult, therefore, we use these shallow frames when looking at the surface brightness profiles, but use only the aligned deep co-add when analyzing the stellar kinematics and stellar populations. Twilight flat field frames, bias frames, and wavelength calibration Hg and Cd lamps were taken during the day.  Flux standard stars were observed in twilight.

@@

Once the Vaccine routines are completed we make the following additional reductions using standard IRAF routines and custom IDL scripts.  The reduced twilight frames are used to generate a fiber-to-fiber flux correction (similar to a longslit illumination correction).  We use the b-spline procedure described in \citet{Kelson03} to generate background frames.  The corners contain enough empty fibers to measure the sky background in each exposure.  Because we bin fibers together, imperfect sky-subtraction represents a serious source of potential systematic errors.

@@

By eye, GLSB systems resemble large faint spiral disks at large radii, and regular ellipticals at their centers.  In the SDSS DR10 \citep{Ahn14}, the Galaxy Zoo classifications \citep{Willett13} are spiral for Malin 1 and 2 and UGC 6614.  In this section, we compare the properties of the GLSB centers and disks with other systems.

@@

Both of our objects, as well as Malin 1, have been observed with SDSS, making it possible to compare them to a large number of systems.  The outer disks of GLSBs are faint enough that the SDSS photometric pipeline effectively ignores them.

To make a comparison sample of galaxies from SDSS, we select galaxies with redshifts between 0.01 and 0.07 and de-reddened $r$ magnitude between 13.5 and 17.7.  When comparing to just elliptical galaxies we make the additional color-cut of $u-r > 2.1$ (and some morphology selectors).  These are simply galaxies for comparison, not a rigorously defined volume-complete sample.  Our full SQL queries are shown in \S\ref{SQL}.  This gives us a sample of 141,001 objects, 45,703 of which have red colors.  Throughout, we use the CosmoloPy\footnote{\url{http://roban.github.com/CosmoloPy/}} fiducial cosmology ($h=0.704$, $\Omega_{m_0}=0.2726$, $\Omega_{\lambda_0}=0.7274$).  Note that we are not correcting for Malmquist bias or biases in the SDSS spectroscopic selection criteria.  Our selection is simply for galaxies of a similar magnitude and redshift.

@@

\begin{figure*}
\epsscale{.4}
\plotone{Plots/Color_Magnitu_Diagram.eps}\plotone{Plots/Faber_Jackson_Relation.eps}\\
\plotone{Plots/Fundamental_Plane.eps}\plotone{Plots/mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin 2 and UGC 6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band fundamental plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   Looking at the SDSS data, our GLSB galaxies appear as relatively typical red elliptical galaxies.  \label{elliptical_relations}}
\end{figure*}

We follow \citet{Bernardi03,Bernardi03b} in fitting fundamental plane parameters to our selected SDSS galaxies.  We fit the fundamental plane by minimizing $\chi^2$ in, 
\begin{equation}
\chi^2 = \log R_0 - a \log \sigma - b \log I_0 - c
\end{equation}
where $R_0$ is the galaxy's effective radius, $\sigma$ is velocity dispersion in kilometers per second and $I_0$ is the surface brightness.  The resulting best-fit parameters are  a =1.376 , b = -0.661, c = -9.938.  We note that we have not made a rigorous volume-complete selection of galaxies, rather, simply galaxies with similar magnitudes, colors, and redshifts to our GLSBs.   See \citet{Saulder13} for a more rigorous analysis of the fundamental plane in SDSS.

@@

The diameters of GLSB disks are very large, with Malin~1's stellar disk detected out to a radius of 80 kpc \citep{Galaz15}.  Stellar disks are typically embedded in much larger \HI\ disks.  The size of the GLSB stellar disks is similar to the largest \HI\ disks observed in nearby galaxies, in support of the \citet{Mapelli08} model where the GLSB disks are the faded remnants of a normal \HI\ disk that has experienced a merger induced star-burst.

@@

In Figure~\ref{disk_comp}, we compare the sizes of the stellar disks of the GLSB galaxies with other observations.  For this plot, we take the maximum detected radius from the literature for the galaxies.  It is possible that deeper observations could increase the values of $R_{max}$.  While UGC 6614 is close in size to the upper end of local \HI\ disks, Malin 1 and 2 are much larger than galaxies observed in the THINGS survey.  This seems to point against the GLSB formation mechanism of \citet{Mapelli08}, as we do not observe \HI\ disks large enough to convert into GLSB systems unless an interaction is able to boost the \HI\ angular momentum.

% /Users/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure*}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, gLSB \HI\ data from Malin 1 from \citet{Pickering97} and gLSB stellar radii from \citet{Galaz15}, \citet{Kasparova14}, and this paper.  Here, $R_{max}$ is the maximum radius with an \HI\ or optical detection.  For the gLSB galaxies, the solid points show the maximum \HI\ detection radius and stars-shaped points show the maximum detected stellar light radius.  These are all maximum detection radii, and thus should be considered lower limits.   \label{disk_comp}}
\end{figure*}

@@

As we have shown in \S\ref{sec:centers} and \S\ref{sec:disks}, GLSB galaxies can be well described as large normal ellipticals that have managed to accrete a high angular momentum disk of stars and gas.  Given the total \HI mass in GLSBs, the accreted object is consistent with a much lower-mass gas-rich spiral.  If this is in fact the way GLSBs form, the disk kinematics represent a unique probe of the dark matter halo of elliptical galaxies.

@@

The Malin 2 rotation curve (Figure~\ref{malin_kinem}) is rather complicated, having several abrupt rises.  This is especially strange given the very smooth photometric profile.  UGC 6614 (Figure~\ref{ugc_kinem}), on the other hand, has a relatively normal spiral galaxy-like rotation curve with a linear rise and flat outer region.  One important difference is that the linear rise extends for 10 kpc.

@@

@@

@@

\begin{figure*}
\epsscale{0.25}
\plotone{Plots/dataMalin2.eps}\plotone{Plots/modelMalin2.eps}\plotone{Plots/data_modelMalin2.eps} \\
\plotone{Plots/databinned_Malin2.eps}\plotone{Plots/modelbinned_Malin2.eps}\plotone{Plots/data_modelbinned_Malin2.eps}
%\epsscale{.5}
\epsscale{1}
\plottwo{Plots/kinembinned_Malin2.eps}{Plots/kinem_HI_binned_Malin2.eps}
\caption{Fitting the line-of-sight velocity maps measured with Ppxf with the kinemetry package.  The dashed curve shows the expected stellar mass contribution to the rotation curve.  Vertical lines mark $R_e$ and $5R_e$.  \HI\ data from \citet{Pickering97}.  \label{malin_kinem}}
\end{figure*}

@@

\begin{figure*}
\epsscale{0.25}
\plotone{Plots/dataUGC06614.eps}\plotone{Plots/modelUGC06614.eps}\plotone{Plots/data_modelUGC06614.eps} \\
\plotone{Plots/databinned_UGC06614.eps}\plotone{Plots/modelbinned_UGC06614.eps}\plotone{Plots/data_modelbinned_UGC06614.eps}
%\epsscale{.5}
\epsscale{1}
\plottwo{Plots/kinembinned_UGC06614.eps}{Plots/kinem_HI_binned_UGC06614.eps}
\caption{Fitting the velocity maps with kinemetry package. Vertical lines mark $R_e$ and $5R_e$.  The dashed lines shows a rotation curve based in the enclosed flux, normalized so that all the SDSS stellar mass is contained within $3R_e$.  \HI\ data from \citet{Pickering97}.\label{ugc_kinem}}
\end{figure*}

>>>>

\altaffiltext{1}{Department of Astronomy, University of Washington, Box 351580,
Seattle WA, 98195}
\altaffiltext{2}{University of Michigan}
\altaffiltext{3}{Jeremiah Horrocks Institute, University of Central Lancashire, Preston PR1 2HE, UK.  RCUK Fellow}
\altaffiltext{4}{California Institute of Technology}

\begin{abstract}
We present IFU observations of the giant low surface brightness galaxies Malin~2 and UGC~6614.  We fit the stellar dynamics of these systems and compare their spatially resolved spectra to star formation history models.  We argue that Malin~2 and UGC~6614 are normal elliptical galaxies which have undergone high angular momentum mergers which formed their GLSB disks, making them unique systems for probing the dark matter halos of elliptical systems.  We find the stellar dynamics are consistent with earlier measurements of the GLSB \HI\ disks.  The stellar populations of the GLSB disks are dominated by old metal-poor stars, while the central ellipticals are metal rich.

@@

Observers tend to classify GLSB systems as normal spirals.  \citet{Matthews01b} list Malin~2 as an Scd while \citet{Das13} concludes that GLSB systems are an extreme form of late type spiral galaxies.  Similarly, \citet{Kasparova14} claim that the large dark matter scale of Malin~2 can explain it's features.  Even the Galaxy Zoo project \citep{Willett13}  classifies GLSB galaxies as spirals.  While GLSB systems superficially resemble normal spiral galaxies, they are clearly extreme outliers when their scale lengths are compared to other galaxy samples, being around a factor of 2 larger than the largest typical spirals \citep{Dalcanton97}.

The disks in GLSB systems are incredibly large, with Malin~1 being detected in $R$-band to a radius of over 100 kpc \citep{Moore06}.  Disks of this size are a challenge for typical $\Lambda$CDM cosmology galaxy formation and exotic formation scenarios have been proposed for GLSB systems.  \citet{Mapelli08} form GLSB-like galaxies in simulations via the fading of collisional ring galaxies.  In their model, a collision sets off an expanding ring of star formation.  As the star formation subsides and the stellar population ages, the ring galaxy fades over $\sim$100 Myr to resembles observed GLSBs.  One obvious problem with this scenario is that most GLSB's are relatively isolated, and thus there are no obvious candidates for the impactor systems that would have triggered the star formation burst.

The other popular formation scenario for GLSB systems is that they are normal galaxies that undergo a high angular momentum minor merger where the LSB disk is the shredded remains of an accreted galaxy.  \citet{Pen06} find that extended disks can be formed via the disruption of dwarf galaxies.  These simulated debris disks have exponential profiles and extend out to 200 kpc from the central galaxy.

@@

In comparison, the stellar disk of Malin~2 extends beyond 40 kpc (5.9 $R_e$), and the HI disk reaches beyond 100 kpc (14.8 $R_e$).  \citet{Galaz15} take deep images of Malin~1 and find the stellar disk diameter extends to 160 kpc.

%These distances have occasionally been reached by x-ray observations of giant elliptical galaxies \citep{Buote12}.

@@

In one of the most extensive studies of a GLSB system, \citet{Kasparova14} combine ARC 3.5m photometry and long-slit spectra with Gemini spectra to mass-model Malin~2.  They find the broadband colors of the extended disk are not consistent with a catastrophic formation, but are best fit by an exponentially declining SFH.  They speculate the main difference between Malin~2 and non-giant LSB galaxies is the dark matter halo scale.  To some extent, this is not a particularly satisfying conclusion because one is then left with the mystery of why Malin~2's dark matter halo is so different from non-giant galaxies.

XXX-in this paper we blah dee blah. In \S\ref{sec:obs} blah blah.

@@

\section{Observations}\label{sec:obs}

@@

For each galaxy, observations were taken at three unique dither positions to provide nearly complete spatial coverage.  Individual science exposures were 30 minutes.  Malin~2 had 10 exposures per dither position (11 exposures for dither position 1), for a total of 15 hours of exposure time.  Observations of UGC 06614 had alignment issues between observing runs.  This resulted in a deep stack of 6 exposures, and a shallow stack with only 2 exposures.  With only 2 exposures, cosmic ray rejection becomes unreliable, therefore, we use these shallow frames when looking at the surface brightness profiles, but use only the aligned deep co-add when analyzing the stellar kinematics and stellar populations. Twilight flat field frames, bias frames, and wavelength calibration Hg and Cd lamps were taken during the day.  Flux standard stars were observed in twilight.

@@

Once the Vaccine routines are completed we make the following additional reductions using standard IRAF routines and custom IDL scripts.  The reduced twilight frames are used to generate a fiber-to-fiber flux correction (similar to a longslit illumination correction).  We use the b-spline procedure described in \citet{Kelson03} to generate background frames.  The corners of the IFU array contain enough empty fibers to measure the sky background in each exposure.  Because we bin fibers together, imperfect sky-subtraction represents a serious source of potential systematic errors.

@@

By eye, GLSB systems resemble large faint spiral disks at large radii, and regular ellipticals at their centers.  In the Sloan Digital Sky Survey Data Relase 10 (SDSS DR10) \citep{Ahn14}, the Galaxy Zoo classifications \citep{Willett13} are spiral for Malin 1 and 2 and UGC 6614.  In this section, we compare the properties of the GLSB centers and disks with other systems.

@@

Both of our objects, as well as Malin~1, have been observed with SDSS, making it possible to compare them to a large number of systems.  The outer disks of GLSBs are faint enough that the SDSS photometric pipeline effectively ignores them.

To make a comparison sample of galaxies from SDSS, we select galaxies with redshifts between 0.01 and 0.07 and de-reddened $r$ magnitude between 13.5 and 17.7.  When comparing to just elliptical galaxies we make the additional color-cut of $u-r > 2.1$ (and some morphology cuts).  These are simply galaxies for comparison, not a rigorously defined volume-complete sample.  Our full SQL queries are shown in \S\ref{SQL}.  This gives us a sample of 141,001 objects, 45,703 of which have red colors.  Throughout, we use the CosmoloPy\footnote{\url{http://roban.github.com/CosmoloPy/}} fiducial cosmology ($h=0.704$, $\Omega_{m_0}=0.2726$, $\Omega_{\lambda_0}=0.7274$).  Note that we are not correcting for Malmquist bias or biases in the SDSS spectroscopic selection criteria.  Our selection is simply for galaxies of a similar magnitude and redshift.

@@

\begin{figure*}
\epsscale{.4}
\plotone{Plots/Color_Magnitu_Diagram.eps}\plotone{Plots/Faber_Jackson_Relation.eps}\\
\plotone{Plots/Fundamental_Plane.eps}\plotone{Plots/mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin~2 and UGC~6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band fundamental plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   From the point of view of the SDSS data reduction pipeline, GLSB galaxies appear as relatively normal red elliptical galaxies.  \label{elliptical_relations}}
\end{figure*}

We follow \citet{Bernardi03,Bernardi03b} in fitting fundamental plane parameters to our selected SDSS galaxies.  We fit the fundamental plane by minimizing $\chi^2$ in, 
\begin{equation}
\chi^2 = \log R_0 - a \log \sigma - b \log I_0 - c
\end{equation}
where $R_0$ is the galaxy's effective radius, $\sigma$ is velocity dispersion in kilometers per second and $I_0$ is the surface brightness.  The resulting best-fit parameters are  $a =1.376$ , $b = -0.661$, $c = -9.938$.  We note that we have not made a rigorous volume-complete selection of galaxies, rather, simply selected galaxies with similar magnitudes, colors, and redshifts to our GLSBs.   See \citet{Saulder13} for a more rigorous analysis of the fundamental plane in SDSS.

@@

XXX--do find-replace check for Malin 2 and UGC 6614 to Malin~2 and UGC~6614

@@

The diameters of GLSB disks are very large, with Malin~1's stellar disk detected out to a radius of 80 kpc \citep{Galaz15}.  Stellar disks are typically embedded in much larger \HI\ disks.  In Figure~\ref{disk_comp}, we compare the sizes of the stellar and \HI\ disks of the GLSB galaxies with other observations.  For this plot, we take the maximum detected radius from the literature for the galaxies.  It is possible that deeper observations could increase the values of $R_{max}$.  While the stellar disk of UGC~6614 is close in size to local observed \HI\ disks, Malin~1 and~2 are larger than hte \HI\ disks observed in the THINGS survey.  This seems to point against the GLSB formation mechanism of \citet{Mapelli08}, as we do not observe \HI\ disks in normal spirals large enough to convert into GLSB systems unless an interaction is able to also boost the angular momentum of the resulting stellar disk.

% /Users/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure*}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, GLSB \HI\ data from Malin 1 from \citet{Pickering97} and GLSB stellar radii from \citet{Galaz15}, \citet{Kasparova14}, and this paper.  Here, $R_{max}$ is the maximum radius with an \HI\ or optical detection.  For the GLSB galaxies, the solid points show the maximum \HI\ detection radius and stars-shaped points show the maximum detected stellar light radius.  All points are maximum detection radii, and thus should be considered lower limits.   \label{disk_comp}}
\end{figure*}

@@

As we have shown in \S\ref{sec:centers} and \S\ref{sec:disks}, GLSB galaxies can be well described as large normal ellipticals that have managed to accrete a high angular momentum disk of stars and gas.  Given the total \HI\ mass in GLSBs, the accreted object is consistent with a much lower-mass gas-rich spiral.  If this is in fact the way GLSBs form, the disk kinematics represent a unique probe of the dark matter halo of elliptical galaxies.

@@

The Malin~2 rotation curve (Figure~\ref{malin_kinem}) is rather complicated, having several abrupt rises.  This is especially strange given the very smooth photometric profile.  UGC~6614 (Figure~\ref{ugc_kinem}), on the other hand, has a relatively normal spiral galaxy-like rotation curve with a linear rise and flat outer region.  One important difference is that the linear rise extends for 10 kpc.

@@

\citet{Wilkinson15} find -0.15 dex/$R_e$ gradients for early type galaxies and flat metallitcity gradients for late type galaxies.

@@

\citet{Hagen16} present the discovery of a GLSB system around UGC~1382.  As in this work, they find the best explination of the extended disk as being the remnants of an accreted drawf galaxy similar to the simulations of \citet{Pen06}.

@@

\begin{figure*}
\epsscale{0.25}
\plotone{Plots/dataMalin2.eps}\plotone{Plots/modelMalin2.eps}\plotone{Plots/data_modelMalin2.eps} \\
\plotone{Plots/databinned_Malin2.eps}\plotone{Plots/modelbinned_Malin2.eps}\plotone{Plots/data_modelbinned_Malin2.eps}
%\epsscale{.5}
\epsscale{1}
\plottwo{Plots/kinembinned_Malin2.eps}{Plots/kinem_HI_binned_Malin2.eps}
\caption{XXX--break this into 2 figures!  Fitting the line-of-sight velocity maps measured with Ppxf with the kinemetry package.  The dashed curve shows the expected stellar mass contribution to the rotation curve.  Vertical lines mark $R_e$ and $5R_e$.  \HI\ data from \citet{Pickering97}.  \label{malin_kinem}}
\end{figure*}

@@

\begin{figure*}
\epsscale{0.25}
\plotone{Plots/dataUGC06614.eps}\plotone{Plots/modelUGC06614.eps}\plotone{Plots/data_modelUGC06614.eps} \\
\plotone{Plots/databinned_UGC06614.eps}\plotone{Plots/modelbinned_UGC06614.eps}\plotone{Plots/data_modelbinned_UGC06614.eps}
%\epsscale{.5}
\epsscale{1}
\plottwo{Plots/kinembinned_UGC06614.eps}{Plots/kinem_HI_binned_UGC06614.eps}
\caption{XXX-break into 2 figures! Fitting the velocity maps with kinemetry package. Vertical lines mark $R_e$ and $5R_e$.  The dashed lines shows a rotation curve based in the enclosed flux, normalized so that all the SDSS stellar mass is contained within $3R_e$.  \HI\ data from \citet{Pickering97}.\label{ugc_kinem}}
\end{figure*}
